Practical Optional Types for Clojure

### 2015-07-10 formal-model.tex
rewrite 3.1

\paragraph{Typing rules}
The core type rules are
given as \figref{main:figure:othertypingrules}. We introduce
the interesting rules while demonstrating how to check
the complement number predicate
\begin{equation}
\abs{\makelocal{d}}{\Top}{\ifexp{\appexp{\numberhuh{}}{\makelocal{d}}}{\false{}}{\true{}}}
\end{equation}
%, including
%a subsumption rule T-Subsume and rules for the false values---T-Nil and T-False---encoded 
%as impossible (\botprop{}) `then' propositions.

The lambda rule T-Abs introduces \isprop{\Top}{\makelocal{d}}
to check the body.
With \propenv{} = \isprop{\Top}{\makelocal{d}},
T-If then checks the test {\appexp{\numberhuh{}}{\makelocal{d}}}
with T-App with three main steps.

First, the operator \numberhuh{} is checked with T-Const, which
uses 
the \constanttypeliteral{} relation to give types to constants.
\numberhuh{} is a predicate over numbers, and
\classconst{} returns its argument's class.
$$
\constanttypefigure{}
$$
The dynamic semantics for constants are given as \figref{main:figure:primitivesem}.

Resuming {\appexp{\numberhuh{}}{\makelocal{d}}},
the operand \makelocal{d} is checked with
T-Local as
$$
\judgementselfrewrite{\propenv{}}
                     {\makelocal{d}}
                     {\UnionNilNum}
                     {\filterset{\notprop{\falsy}{\makelocal{d}}}
                                {\isprop{\falsy}{\makelocal{d}}}}
                     {\makelocal{d}}
$$
which encodes the type, proposition, and object information
previously discussed about variables.

Finally, T-App substitutes the operand's object for the parameter
---here \makelocal{d} for \numberhuh{}'s \x{}---yielding
$$
\judgementselfrewrite{\isprop{\Top}{\makelocal{d}}}{\appexp{\numberhuh{}}{\makelocal{d}}}
  {\Boolean{}}
  {\filterset{\isprop{\Number{}}{\makelocal{d}}}
             {\notprop{\Number{}}{\makelocal{d}}}}
  {\emptyobject{}}
$$

To check {\ifexp{\appexp{\numberhuh{}}{\makelocal{d}}}{\false{}}{\true{}}},
T-If then
introduces \isprop{\Number{}}{\makelocal{d}}
in the `then' branch and \notprop{\Number{}}{\makelocal{d}}
in the `else' branch.
%The object \x{} over latent propositions \thenprop{\prop{}} and
%\elseprop{\prop{}}, latent object \object{}, and latent type \t{}. The
%actual argument object---\objectp{}---is substituted in at application by T-App.

The T-App rule instantiates parameters---like \x{} in T-Abs---with their actual object.
The expression {\appexp{\numberhuh{}}{\makelocal{d}}} replaces its parameter object with \
\judgementselfrewrite{\isprop{\Top}{\makelocal{d}}}{\appexp{\numberhuh{}}{\makelocal{d}}}
  {\Boolean{}}
  {\filterset{\isprop{\Number{}}{\makelocal{d}}}
             {\notprop{\Number{}}{\makelocal{d}}}}
  {\emptyobject{}}

The T-If refines each branch based on the test---a
{\appexp{\numberhuh{}}{\makelocal{d}}} test introduces \isprop{\Number{}}{\makelocal{d}}
for checking the `then' branch and \notprop{\Number{}}{\makelocal{d}}
for checking the `else' branch.
Information from each branch is combined via subsumption for the overall type, propositions, and object.

For example
$
\judgementselfrewrite{\isprop{\Top}{\makelocal{d}}}
  {\ifexp{\appexp{\numberhuh{}}{\makelocal{d}}}{\false{}}{\true{}}}
  {\Boolean{}}
  {\filterset{\notprop{\Number{}}{\makelocal{d}}}
             {\isprop{\Number{}}{\makelocal{d}}}}
  {\emptyobject{}}
$
type checks because T-Subsume allows us to check both branches as
$
\judgementselfrewrite{\isprop{\Number{}}{\makelocal{d}}}
  {\false{}}
  {\Boolean{}}
  {\filterset{\notprop{\Number{}}{\makelocal{d}}}
             {\isprop{\Number{}}{\makelocal{d}}}}
  {\emptyobject{}}
$
and
$
\judgementselfrewrite{\notprop{\Number{}}{\makelocal{d}}}
  {\true{}}
  {\Boolean{}}
  {\filterset{\notprop{\Number{}}{\makelocal{d}}}
             {\isprop{\Number{}}{\makelocal{d}}}}
  {\emptyobject{}}
$
respectively.
%
%For example, if \e{2} is true when variable {\makelocal{y}} is a \File{}
%and \e{3} is true when variable {\makelocal{z}} is a \Number{}, then
%we use T-Subsume on both branches
%to introduce a logical disjunction
%since at least one must be true if the entire expression is true.

@@

The let rule T-Let links inferred information about
\x{} to the expression used to instantiate \x{}, \ep{1}, via logical implications.

%The T-Local rule connects the type system to the proof system over type propositions
%via \inpropenv {\propenv{}} {\isprop {\t{}} {\x{}}}
%to derive a type for a variable.
%Using this rule, the type system can then appeal to L-Update to refine the type
%assigned to \x{}.
%
%We are now equipped to type check
%\egref{example:parent-if}:
%$$
%\clj{(if f (.getParent f) nil)}
%$$
%
%With {\propenv{}} = {\isprop{\UnionNilFile{}}{f}},
%$$
%\judgement{\propenv{}}{f}{\UnionNilFile{}}{\localfilterset{f}}{f}
%$$
%via T-Local.
%
%Checking the then branch involves extending
%the proposition environment with {\notprop{\falsy}{f}}
%$$
%\judgement{{\propenv{}},{\isprop{\Number}{\x{}}}}{\x{}}{\Number{}}{\filterset{\notprop{\falsy{}}{\x{}}}{\isprop{\falsy{}}{\x{}}}}{\emptyobject{}}
%$$
%because we can now satisfy the premise of T-Local:
%$$
%\inpropenv{{\propenv{}},\isprop{\Number}{\x{}}}{\isprop{\Number}{\x{}}}.
%$$
%\judgement{{\propenv{}},\isprop{\Number}{\x{}}}{\hastype{\appexp{\inc{}}{\x{}}}{\Number{}}}{\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
%$$
%$$
%\judgement{{\propenv{}},\notprop{\Number}{\x{}}}{\hastype{\zeroliteral{}}{\Number}}{\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
%$$

@@

\paragraph{Constructors} First we check and convert {\newexp {\File{}} {\makestr{a/b}}} to {\newstaticexp {\String} {\File{}} {\File{}} {\makestr{a/b}}}
The T-New typing rule checks and rewrites constructors.
$$
    {\TNew}
$$
To check
{\newexp {\File{}} {\makestr{a/b}}}
we first resolve the constructor overload in the class table---there is at most one
to simplify presentation.
With \classhint{1} = \String,
we convert to a nilable type the argument with \t{1} = \Union{\Nil}{\String}
and type check {\makestr{a/b}} against \t{1}.
Typed Clojure defaults to allowing non-nilable arguments, but this
can be overridden, so we model the more general case.
% which erases nil
The return Java type \File is converted to a non-nil
Typed Clojure type \t{} = \File for the return type,
and the propositions say constructors can never be false---constructors
can never produce the internal boolean value that Clojure uses for \false{}, or \nil{}.
Finally, the constructor rewrites to {\newstaticexp {\String} {\File{}} {\File{}} {\makestr{a/b}}}.

@@

For convenience, examples in this section are flattened when they are really nested
let bindings. We also elide trivial latent propositions and objects.
The following is an abbreviation of the previous expression.
\begin{lstlisting}
${\createmultiexp {\ArrowTwo{\x{}}{\Keyword}{\String}} {\abs{\x{}}{\Keyword}{\x{}}}}$
${\extendmultiexp {hi} {\makekw{en}} {\abs {\x{}} {\Keyword} {\makestr{hello}}}}$
${\extendmultiexp {hi} {\makekw{fr}} {\abs {\x{}} {\Keyword} {\makestr{bonjour}}}}$
$\appexp{hi}{\makekw{en}}$
\end{lstlisting}

>>>>

\paragraph{Typing rules}
The core typing rules are
given as \figref{main:figure:othertypingrules}. We introduce
the interesting rules with the complement number predicate
as a running example.
\begin{equation}
\abs{\makelocal{d}}{\Top}{\ifexp{\appexp{\numberhuh{}}{\makelocal{d}}}{\false{}}{\true{}}}
\end{equation}
%, including
%a subsumption rule T-Subsume and rules for the false values---T-Nil and T-False---encoded 
%as impossible (\botprop{}) `then' propositions.

The lambda rule T-Abs introduces \isprop{\s{}}{\x{}}} = \isprop{\Top}{\makelocal{d}}
to check the body.
With \propenv{} = \isprop{\Top}{\makelocal{d}},
the T-If rule first checks the test \e{1} = {\appexp{\numberhuh{}}{\makelocal{d}}}
via the T-App rule, with three main steps.

First, in T-App the operator \e{} = \numberhuh{} is checked with T-Const, which
uses 
the \constanttypeliteral{} relation to give types to constants.
\numberhuh{} is a predicate over numbers, and
\classconst{} returns its argument's class.
$$
\constanttypefigure{}
$$
The dynamic semantics for constants are given as \figref{main:figure:primitivesem}.

Resuming {\appexp{\numberhuh{}}{\makelocal{d}}},
in T-App the operand \ep{} = \makelocal{d} is checked with
T-Local as
\begin{equation}
\judgementselfrewrite{\propenv{}}
                     {\makelocal{d}}
                     {\Top}
                     {\filterset{\notprop{\falsy}{\makelocal{d}}}
                                {\isprop{\falsy}{\makelocal{d}}}}
                     {\makelocal{d}}
\end{equation}
which encodes the type, proposition, and object information
about variables
as previously discussed.

Finally, the T-App rule substitutes the operand's object \objectp{}
for the parameter \x{} in the latent type, propositions, and object.
\begin{equation}
\judgementselfrewrite{\propenv{}}
  {\appexp{\numberhuh{}}{\makelocal{d}}}
  {\Boolean{}}
  {\filterset{\isprop{\Number{}}{\makelocal{d}}}
             {\notprop{\Number{}}{\makelocal{d}}}}
  {\emptyobject{}}
\end{equation}
To demonstrate, the `then' proposition---in T-App {\replacefor {\thenprop{\prop{}}} {\objectp{}} {\x{}}}---substitutes
the latent `then' proposition of \constanttype{\numberhuh{}} with 
\makelocal{d}, giving
{\replacefor {\isprop{\Number{}}{\x{}}} {\makelocal{d}} {\x{}}} =
{\isprop{\Number{}}{\makelocal{d}}}.

To check the branches of {\ifexp{\appexp{\numberhuh{}}{\makelocal{d}}}{\false{}}{\true{}}},
T-If
introduces \thenprop{\prop{1}} = \isprop{\Number{}}{\makelocal{d}}
in the `then' branch 
\e{2} = {\false{}},
and \elseprop{\prop{1}} = \notprop{\Number{}}{\makelocal{d}}
in the `else' branch
\e{3} = \true{}.

The branches are checked with T-False and T-True respectively,
%$$
%\judgementselfrewrite{\propenv{},{\isprop{\Number{}}{\makelocal{d}}}}
%  {\false{}}
%  {\False{}}
%  {\filterset{\botprop{}}
%             {\topprop{}}}
%  {\emptyobject{}}
%$$
then T-Subsume.
$$
\begin{array}{c}
\judgementselfrewrite{\propenv{},{\isprop{\Number{}}{\makelocal{d}}}}
  {\false{}}
  {\Boolean{}}
  {\filterset{\notprop{\Number{}}{\makelocal{d}}}
             {\isprop{\Number{}}{\makelocal{d}}}}
  {\emptyobject{}}
  \\
\judgementselfrewrite{\propenv{},{\notprop{\Number{}}{\makelocal{d}}}}
  {\true{}}
  {\Boolean{}}
  {\filterset{\notprop{\Number{}}{\makelocal{d}}}
             {\isprop{\Number{}}{\makelocal{d}}}}
  {\emptyobject{}}
\end{array}
$$
%to suit the T-If outputs \t{} = \Boolean{}, \thenprop{\prop{}}
%= {\notprop{\Number{}}{\makelocal{d}}}, \elseprop{\prop{}} = {\isprop{\Number{}}{\makelocal{d}}},
%and \object{} = {\emptyobject{}}.
%
%In T-Subsume, we can upcast \t{} = \False{} to \tp{} = \Boolean{} via the premise 
%\issubtypein{}{\t{}}{\tp{}}.
%and 
%\inpropenv {\propenv{}, {\elseprop {\prop{}}}} {\elseprop {\propp{}}}
%from 
%{\elseprop {\prop{}}} = \topprop{} to {\elseprop {\propp{}}} = {\isprop{\Number{}}{\makelocal{d}}}.
The T-Subsume conditions
\inpropenv {\propenv{}, {\thenprop {\prop{}}}} {\thenprop {\propp{}}}
and
\inpropenv {\propenv{}, {\elseprop {\prop{}}}} {\elseprop {\propp{}}}
allow us to pick compatible propositions.

Finally T-Abs concludes, using the T-If outputs, with
$$
{\judgementselfrewrite{}{\abs{\makelocal{d}}{\Top}{\ifexp{\appexp{\numberhuh{}}{\makelocal{d}}}{\false{}}{\true{}}}}
                    {\ArrowOne {\makelocal{d}} {\Top{}}
                                      {\Boolean{}}
                                      {\filterset {\notprop{\Number{}}{\makelocal{d}}}
                                                  {\isprop{\Number{}}{\makelocal{d}}}}
                                      {\emptyobject{}}}
                    {\filterset {\topprop{}}
                                {\botprop{}}}
                    {\emptyobject{}}}
$$

%The object \x{} over latent propositions \thenprop{\prop{}} and
%\elseprop{\prop{}}, latent object \object{}, and latent type \t{}. The
%actual argument object---\objectp{}---is substituted in at application by T-App.

%The T-App rule instantiates parameters---like \x{} in T-Abs---with their actual object.
%The expression {\appexp{\numberhuh{}}{\makelocal{d}}} replaces its parameter object with \
%\judgementselfrewrite{\isprop{\Top}{\makelocal{d}}}{\appexp{\numberhuh{}}{\makelocal{d}}}
%  {\Boolean{}}
%  {\filterset{\isprop{\Number{}}{\makelocal{d}}}
%             {\notprop{\Number{}}{\makelocal{d}}}}
%  {\emptyobject{}}

%The T-If refines each branch based on the test---a
%{\appexp{\numberhuh{}}{\makelocal{d}}} test introduces \isprop{\Number{}}{\makelocal{d}}
%for checking the `then' branch and \notprop{\Number{}}{\makelocal{d}}
%for checking the `else' branch.
%%Information from each branch is combined via subsumption for the overall type, propositions, and object.
%The T-Subsume rule

%For example
%$
%\judgementselfrewrite{\isprop{\Top}{\makelocal{d}}}
%  {\ifexp{\appexp{\numberhuh{}}{\makelocal{d}}}{\false{}}{\true{}}}
%  {\Boolean{}}
%  {\filterset{\notprop{\Number{}}{\makelocal{d}}}
%             {\isprop{\Number{}}{\makelocal{d}}}}
%  {\emptyobject{}}
%$
%type checks because T-Subsume allows us to check both branches as
%$
%\judgementselfrewrite{\isprop{\Number{}}{\makelocal{d}}}
%  {\false{}}
%  {\Boolean{}}
%  {\filterset{\notprop{\Number{}}{\makelocal{d}}}
%             {\isprop{\Number{}}{\makelocal{d}}}}
%  {\emptyobject{}}
%$
%and
%$
%\judgementselfrewrite{\notprop{\Number{}}{\makelocal{d}}}
%  {\true{}}
%  {\Boolean{}}
%  {\filterset{\notprop{\Number{}}{\makelocal{d}}}
%             {\isprop{\Number{}}{\makelocal{d}}}}
%  {\emptyobject{}}
%$
%respectively.
%
%For example, if \e{2} is true when variable {\makelocal{y}} is a \File{}
%and \e{3} is true when variable {\makelocal{z}} is a \Number{}, then
%we use T-Subsume on both branches
%to introduce a logical disjunction
%since at least one must be true if the entire expression is true.

@@

%The let rule T-Let links inferred information about
%\x{} to the expression used to instantiate \x{}, \ep{1}, via logical implications.
%
%The T-Local rule connects the type system to the proof system over type propositions
%via \inpropenv {\propenv{}} {\isprop {\t{}} {\x{}}}
%to derive a type for a variable.
%Using this rule, the type system can then appeal to L-Update to refine the type
%assigned to \x{}.
%
%We are now equipped to type check
%\egref{example:parent-if}:
%$$
%\clj{(if f (.getParent f) nil)}
%$$
%
%With {\propenv{}} = {\isprop{\UnionNilFile{}}{f}},
%$$
%\judgement{\propenv{}}{f}{\UnionNilFile{}}{\localfilterset{f}}{f}
%$$
%via T-Local.
%
%Checking the then branch involves extending
%the proposition environment with {\notprop{\falsy}{f}}
%$$
%\judgement{{\propenv{}},{\isprop{\Number}{\x{}}}}{\x{}}{\Number{}}{\filterset{\notprop{\falsy{}}{\x{}}}{\isprop{\falsy{}}{\x{}}}}{\emptyobject{}}
%$$
%because we can now satisfy the premise of T-Local:
%$$
%\inpropenv{{\propenv{}},\isprop{\Number}{\x{}}}{\isprop{\Number}{\x{}}}.
%$$
%\judgement{{\propenv{}},\isprop{\Number}{\x{}}}{\hastype{\appexp{\inc{}}{\x{}}}{\Number{}}}{\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
%$$
%$$
%\judgement{{\propenv{}},\notprop{\Number}{\x{}}}{\hastype{\zeroliteral{}}{\Number}}{\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
%$$

@@

\paragraph{Constructors} First we check and convert {\newexp {\File{}} {\makestr{a/b}}} to {\newstaticexp {\String} {\File{}} {\File{}} {\makestr{a/b}}}.
The T-New typing rule checks and rewrites constructors.
$$
    {\TNew}
$$
To check
{\newexp {\File{}} {\makestr{a/b}}}
we first resolve the constructor overload in the class table---there is at most one
to simplify presentation.
With \classhint{1} = \String,
we convert to a nilable type the argument with \t{1} = \Union{\Nil}{\String}
and type check {\makestr{a/b}} against \t{1}.
Typed Clojure defaults to allowing non-nilable arguments, but this
can be overridden, so we model the more general case.
% which erases nil
The return Java type \File is converted to a non-nil
Typed Clojure type \t{} = \File for the return type,
and the propositions say constructors can never be false---constructors
can never produce the internal boolean value that Clojure uses for \false{}, or \nil{}.
Finally, the constructor rewrites to {\newstaticexp {\String} {\File{}} {\File{}} {\makestr{a/b}}}.

@@

%For convenience, examples in this section are flattened when they are really nested
%let bindings. We also elide trivial latent propositions and objects.
%The following is an abbreviation of the previous expression.
%\begin{lstlisting}
%${\createmultiexp {\ArrowTwo{\x{}}{\Keyword}{\String}} {\abs{\x{}}{\Keyword}{\x{}}}}$
%${\extendmultiexp {hi} {\makekw{en}} {\abs {\x{}} {\Keyword} {\makestr{hello}}}}$
%${\extendmultiexp {hi} {\makekw{fr}} {\abs {\x{}} {\Keyword} {\makestr{bonjour}}}}$
%$\appexp{hi}{\makekw{en}}$
%\end{lstlisting}
